---
title: Data analysis
subtitle: Oilbird vocalizations
author: Marcelo Araya-Salas
date: "`r Sys.Date()`"
toc: true
toc-depth: 3
toc-location: left
number-sections: true
highlight-style: pygments
format:
  html:
    code-fold: true
    code-tools: true
    code-copy: true
    embed-resources: true
editor_options: 
  chunk_output_type: console
---


&nbsp; 


<!-- this code add line numbers to code blocks -->
<!-- only works when code folding is not used in yaml (code_folding: show) -->

```{=html}
<style>
body
  { counter-reset: source-line 0; }
pre.numberSource code
  { counter-reset: none; }
</style>
```

```{r add link to github repo, echo = FALSE, results='asis'}

# print link to github repo if any
if (file.exists("./.git/config")){
  config <- readLines("./.git/config")
  url <- grep("url",  config, value = TRUE)
  url <- gsub("\\turl = |.git$", "", url)[2]
  cat("\nSource code and data found at [", url, "](", url, ")", sep = "")
  }

```

```{r setup style, echo = FALSE, message = FALSE, warning=FALSE}

# options to customize chunk outputs
knitr::opts_chunk$set(
  class.source = "numberLines lineAnchors", # for code line numbers
  tidy.opts = list(width.cutoff = 65), 
  tidy = TRUE,
  message = FALSE,
  warning = FALSE
 )

knitr::opts_knit$set(root.dir = "..")

```


# Load packages {.unnumbered .unlisted}
```{r load packages}

# load function from sketchy
source("https://raw.githubusercontent.com/maRce10/sketchy/main/R/load_packages.R")

# install/ load packages
sketchy::load_packages(packages = c(github = "maRce10/Rraven", github = "maRce10/ohun", github = "maRce10/warbleR", "viridis", "caret", "randomForest", "umap", "mclust"))

```

# Reading annotations
```{r}
clps <- "./DiegoMejia_Grabaciones/5_min_clips/"
spctrs <- "./DiegoMejia_Grabaciones/spectros/"

pth2 <- "./DiegoMejia_Grabaciones/Oilbird_Anotations Diego Mejía"

anns <- imp_raven(pth2, recursive = TRUE, warbler.format = TRUE, name.from.file = TRUE,ext.case = "lower", all.data = TRUE)


empty_recs <- .Options$Rraven$`empty selection table files` 

cs <- check_sels(anns, path = clps)

anns <- anns[cs$check.res == "OK", ]

feature_reference(anns, units = c("s", "kHz"))

anns2 <- anns[anns$Adulto == "y" & anns$Sobrelapamiento == "n", ]

write.csv(anns2, "./data/processed/adult_annotations.csv", row.names = FALSE)
```

# Print spectrograms
```{r, eval = FALSE}

anns <- read.csv("./data/processed/adult_annotations.csv")

spectrograms(anns, flim = c(0, 11), collevels = seq(-120, 0, 5), pal =  viridis::viridis, path = clps, dest.path = spctrs, propwidth = TRUE)

```

# Measure acoustic features
```{r, eval = FALSE}
anns <- read.csv("./data/processed/adult_annotations.csv")

anns$type <- NA
dirs <- list.dirs(path = "./DiegoMejia_Grabaciones/Espectros Categorizados")[-1]

for(x in dirs)
anns$type[filter_sels(anns, path = x, index = TRUE)] <- basename(x)

table(anns$type)

check_sels(anns, path = clps)


acoust_feat <- spectro_analysis(anns, parallel = 20, path = clps)
mfcs <- mfcc_stats(anns, parallel = 20, path = clps)

acoust_feat <- cbind(acoust_feat, mfcs[, -c(1, 2)])

acoust_feat$type <- anns$type

write.csv(acoust_feat, "./data/processed/acoustic_features.csv", row.names = FALSE)

```

# Random Forest classification
```{r, eval = FALSE}

acoust_feat <- read.csv("./data/processed/acoustic_features.csv")

acoust_feat <- acoust_feat[complete.cases(acoust_feat), ]

# acoust_feat <- acoust_feat[!is.na(acoust_feat$type), ]
acoust_feat$type <- make.names(acoust_feat$type)
acoust_feat$type <- factor(acoust_feat$type)

# Create data subsets
partition <- createDataPartition(
    y = acoust_feat$type,
    times = 1,
    p = 0.75,
    list = FALSE
)

trainset <- acoust_feat[partition, -c(1, 2)]
testset <- acoust_feat[-partition, -c(1, 2)]

trcontrol <-
    trainControl(
        method = "repeatedcv",
        number = 100,
        savePredictions = TRUE,
        classProbs = TRUE,
        returnResamp = "all",
        sampling = "down"
    )

pred_model <-
    train(
        type ~ .,
        data = trainset,
        method = "rf",
        trControl = trcontrol,
        metric = "Accuracy",
        preProcess = "scale", 
        proximity = TRUE
    )

# save confusion matrix
conf_mat <-
    confusionMatrix(predict(pred_model, testset), testset$type)

conf_df <- as.data.frame(conf_mat$table)

conf_df$total <-
    sapply(conf_df$Reference, function(x)
        sum(testset$type ==
                x))

conf_df$proportion <- conf_df$Freq / conf_df$total

# fit model on complete data set
best_rf_model <- pred_model$finalModel

all_rf_model <- randomForest(
  type ~ .,
  data = acoust_feat,  # Your entire dataset
  proximity = TRUE,  # Include proximity matrix
  ntree = best_rf_model$ntree,  # Number of trees
  mtry = best_rf_model$mtry,    # Number of variables tried for splitting at each node
  nodesize = best_rf_model$nodesize,  # Minimum size of terminal nodes
  maxnodes = best_rf_model$maxnodes  # Maximum number of terminal nodes
)


saveRDS(
    list(
        pred_model_bb = pred_model,
        conf_mat_bb = conf_mat,
        confusion_df_bb = conf_df,
        testset_bb = testset,
        all_rf_model = all_rf_model,
        data = acoust_feat
    ),
    "./data/processed/random_forest_model_call_types.RDS"
)

```

```{r}

rf_model_results <- readRDS("./data/processed/random_forest_model_call_types.RDS")

# print confusion matrix results
rf_model_results$conf_mat_bb$overall

confusion_df <- rf_model_results$confusion_df_bb

ggplot(confusion_df, aes(x = Reference, y = Prediction, fill = proportion)) +
    geom_tile() + theme_bw() + coord_equal() + scale_fill_distiller(palette = "Greens",
    direction = 1) + geom_text(aes(label = round(proportion, 2)),
    color = "black", size = 3) + theme_classic() + theme(axis.text.x = element_text(color = "black",
    size = 11, angle = 30, vjust = 0.8, hjust = 0.8))


```

# Manual labels

## UMAP visualization
```{r, eval= FALSE}

umap_result <- umap(rf_model_results$all_rf_model$proximity, n_neighbors = 15,
    n_components = 2)

# Create a data frame with the UMAP results
umap_df <- data.frame(UMAP1 = umap_result$layout[, 1], UMAP2 = umap_result$layout[,
    2], type = rf_model_results$data$type)

umap_df$pred.type <- predict(rf_model_results$pred_model_bb, rf_model_results$data)
  # predict(object = rf_model_results$all_rf_model,
    # rf_model_results$data)

mod_umap <- Mclust(umap_df[, 1:2])

summary(mod_umap)

grouping_umap <- as.factor(mod_umap$classification)
umap_df$group <-  mod_umap$classification

write.csv(umap_df, "./data/processed/umap_on_rf_proximity_call_types.csv",
    row.names = FALSE)

```

```{r}
umap_df <- read.csv("./data/processed/umap_on_rf_proximity_call_types.csv")

# Create a scatterplot
gg_umap_loc <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = type,
    fill = type, shape = type)) + geom_point(size = 4) + ylim(c(-7, 7)) + scale_color_viridis_d(alpha = 0.3,
    begin = 0.1, end = 0.8) + scale_fill_viridis_d(alpha = 0.2, begin = 0.1,
    end = 0.8) + theme_classic(base_size = 20) + labs(x = "UMAP1",
    y = "UMAP2", color = "Call type", fill = "Call type", shape = "Call type") + scale_shape_manual(values=rep(c(1:2, 16:21), 3)[1:16])

gg_umap_loc

```


# Unsupervised labels

## UMAP visualization
```{r}

umap_df <- read.csv("./data/processed/umap_on_rf_proximity_call_types.csv")

# Create a scatterplot
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = as.factor(group),
    fill = as.factor(group), shape = as.factor(group))) + geom_point(size = 4) + ylim(c(-7, 7)) + scale_color_viridis_d(alpha = 0.3,
    begin = 0.1, end = 0.8) + scale_fill_viridis_d(alpha = 0.2, begin = 0.1,
    end = 0.8) + theme_classic(base_size = 20) + labs(x = "UMAP1",
    y = "UMAP2", color = "Call type", fill = "Call type", shape = "Call type") + scale_shape_manual(values=rep(c(1:2, 16:22), 3)[1:9])

```

# Match between manual and unsupervised classifications

All calls
```{r}
# confusion matrix
conf_df <- aggregate(UMAP2 ~ type + group, umap_df, length)

names(conf_df)[ncol(conf_df)] <- "Freq"

conf_df$total <-
    sapply(conf_df$group, function(x)
        sum(umap_df$group ==
                x))

conf_df$proportion <- conf_df$Freq / conf_df$total

conf_df <- conf_df[order(conf_df$proportion), ]

conf_df$group <- factor(conf_df$group)

conf_df$type <- factor(conf_df$type)

ggplot(conf_df, aes(y = group, x = type, fill = proportion)) +
    geom_tile() + theme_bw() + coord_equal() + scale_fill_distiller(palette = "Greens",
    direction = 1) + geom_text(aes(label = round(proportion, 2)),
    color = "black", size = 3) + theme_classic() + theme(axis.text.x = element_text(color = "black",
    size = 11, angle = 30, vjust = 0.8, hjust = 0.8)) +
  labs(y = "Unsupervised categories", x = "Manual categories")
```

All calls excluding proportions lower than 5%
```{r}
# confusion matrix
conf_df <- aggregate(UMAP2 ~ type + group, umap_df, length)

names(conf_df)[ncol(conf_df)] <- "Freq"

conf_df$total <-
    sapply(conf_df$group, function(x)
        sum(umap_df$group ==
                x))

conf_df$proportion <- conf_df$Freq / conf_df$total

conf_df <- conf_df[order(conf_df$proportion), ]

conf_df$group <- factor(conf_df$group)

conf_df$type <- factor(conf_df$type)

conf_df <- conf_df[conf_df$proportion > 0.05, ]

ggplot(conf_df, aes(y = group, x = type, fill = proportion)) +
    geom_tile() + theme_bw() + coord_equal() + scale_fill_distiller(palette = "Greens",
    direction = 1) + geom_text(aes(label = round(proportion, 2)),
    color = "black", size = 3) + theme_classic() + theme(axis.text.x = element_text(color = "black",
    size = 11, angle = 30, vjust = 0.8, hjust = 0.8)) +
  labs(y = "Unsupervised categories", x = "Manual categories")
```

Only those that were correctly classified based on manual categories
```{r}
# confusion matrix
conf_df <- aggregate(UMAP2 ~ type + group, umap_df[umap_df$type == umap_df$pred.type,], length)

names(conf_df)[ncol(conf_df)] <- "Freq"

conf_df$total <-
    sapply(conf_df$group, function(x)
        sum(umap_df$group[umap_df$type == umap_df$pred.type] ==
                x))

conf_df$proportion <- conf_df$Freq / conf_df$total

conf_df <- conf_df[order(conf_df$proportion), ]

conf_df$group <- factor(conf_df$group)

conf_df$type <- factor(conf_df$type)

ggplot(conf_df, aes(y = group, x = type, fill = proportion)) +
    geom_tile() + theme_bw() + coord_equal() + scale_fill_distiller(palette = "Greens",
    direction = 1) + geom_text(aes(label = round(proportion, 2)),
    color = "black", size = 3) + theme_classic() + theme(axis.text.x = element_text(color = "black",
    size = 11, angle = 30, vjust = 0.8, hjust = 0.8)) +
  labs(y = "Unsupervised categories", x = "Manual categories")
```

Only those that were correctly classified based on manual categories lower than 5%
```{r}
# confusion matrix
conf_df <- aggregate(UMAP2 ~ type + group, umap_df[umap_df$type == umap_df$pred.type,], length)

names(conf_df)[ncol(conf_df)] <- "Freq"

conf_df$total <-
    sapply(conf_df$group, function(x)
        sum(umap_df$group[umap_df$type == umap_df$pred.type] ==
                x))

conf_df$proportion <- conf_df$Freq / conf_df$total

conf_df <- conf_df[order(conf_df$proportion), ]

conf_df$group <- factor(conf_df$group)

conf_df$type <- factor(conf_df$type)

conf_df <- conf_df[conf_df$proportion > 0.05, ]

ggplot(conf_df, aes(y = group, x = type, fill = proportion)) +
    geom_tile() + theme_bw() + coord_equal() + scale_fill_distiller(palette = "Greens",
    direction = 1) + geom_text(aes(label = round(proportion, 2)),
    color = "black", size = 3) + theme_classic() + theme(axis.text.x = element_text(color = "black",
    size = 11, angle = 30, vjust = 0.8, hjust = 0.8)) +
  labs(y = "Unsupervised categories", x = "Manual categories")
```

## Chi-square on contingency table

```{r}
tbl <- xtabs(Freq ~ group + type, conf_df)

chisq.test(tbl)

```

<div class="alert alert-success">

# Takeaways {.unnumbered .unlisted}

- 

</div>

&nbsp;

<!-- '---' adds a gray vertical line -->

---

&nbsp; 
 
 <!-- add packages used, system details and versions  -->
 
# Session information {.unnumbered .unlisted}

```{r session info, echo=F}

sessionInfo()

```
